---
title: "Class 12: Introduction to Genome Informatics"
author: "Nundini Varshney (PID: A16867985)"
format: html
---

## Section 4: Population Scale Analysis [HOMEWORK]

One sample is obviously not enough to know what is happening in a population. You are 
interested in assessing genetic differences on a population scale. So, you processed about ~230 samples and did the normalization on a genome level. Now, you want to find whether there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3 
expression.

Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes. 

```{r}
#First, I read the file into R

genstats <- read.table("textfile.txt", row.names = 1)
head(genstats)
```
```{r}
# How many rows are there?
nrow(genstats)

# What is the sample size for each genotype?
table(genstats$geno)
```

```{r}
# Find corresponding median levels for each genotype
median <- boxplot(genstat)
```



```{r}
library(ggplot2)
```

#Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

Now, I will create a boxplot for this data. 

```{r}
ggplot(genstats) + aes(geno, exp, fill=geno) +
  geom_boxplot(notch=TRUE)
```

