---
title: "Class 12: Introduction to Genome Informatics"
author: "Nundini Varshney (PID: A16867985)"
format: html
---

## Section 4: Population Scale Analysis [HOMEWORK]

One sample is obviously not enough to know what is happening in a population. You are 
interested in assessing genetic differences on a population scale. So, you processed about ~230 samples and did the normalization on a genome level. Now, you want to find whether there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3 
expression.

> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes. 

```{r}
#First, I read the file into R

genstats <- read.table("textfile.txt", row.names = 1)
head(genstats)
```

Sample size for each genotype?

```{r}
#How many rows are there?
nrow(genstats)

#What is the sample size for each genotype?
table(genstats$geno)
```

Median for each of the three genotypes?

```{r}
genstats <- read.table("textfile.txt", row.names = 1)

# Calculate the median expression level for each genotype
medians <- tapply(genstats$exp, genstats$geno, median)

# Print the median values for each type
print(medians)
```

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative 
expression value between A/A and G/G displayed in this plot? Does the SNP effect the 
expression of ORMDL3?

```{r}
# Generate boxplot
library(ggplot2)
ggplot(genstats, aes(x = geno, y = exp)) +
  geom_boxplot(fill = "lightblue", color = "darkblue") +
  labs(x = "Genotype", y = "Expression",
       title = "Expression Levels by Genotype") 
```
From this boxplot, we can infer that the A/A genotype has a higher expression level than the G/G genotype because visually speaking, the box and its median line is higher in comparison to G/G's. The SNP does affect the expression of ORMDL3.






