---
title: "Class 9: Halloween Candy Mini Project"
author: "Nundini Varshney (PID: A16867985)"
format: html
---

Here we analyze a candy dataset from the 538 website. This is a CSV file from their GitHub repository. 

## Data Import 

```{r}
candy <- read.csv("candy-data.txt", row.names = 1)
head(candy)
```

> Q1.  How many different candy types are in this dataset?

```{r}
nrow(candy)
```

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

## Data Exploration 


> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["M&M", ]$winpercent
```

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```
> Q. What is the least liked candy in the dataset - lowest winpercent 

```{r}
inds <- order(candy$winpercent)
head(candy[inds, ])
```

```{r}
library("skimr")
skim(candy)
skimr::skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Winpercent looks to be on a different scale to the majority of the other columns

> Q7. What do you think a zero and one represent for the candy$chocolate column?

Zero represents if the candy brand is a chocolate whereas a one means it is chocolate. 

> Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent) +
  geom_histogram(binwidth=6)
```

```{r}
hist(candy$winpercent, breaks=8)
```

> Q9. Is the distribution of winpercent values symmetrical?

No, more skewed left

> Q10. Is the center of the distribution above or below 50%?

Below 

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
mean(candy$winpercent[as.logical(candy$chocolate)])
mean(candy$winpercent[as.logical(candy$fruit)])
```
Therefore, chocolate candy on average is raned higher than fruit candy. 

> Q12. Is this difference statistically significant?

```{r}
x <- candy$winpercent[as.logical(candy$chocolate)]
y <- candy$winpercent[as.logical(candy$fruit)]
t.test(x,y)
```
Therefore, it is statistically significant. 

> Q13. What are the five least liked candy types in this set?
> Q14. What are the top 5 all time favorite candy types out of this set?

> Q15. Make a bar plot

```{r}
ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col() +
  labs(x="Win Percent", y=NULL)

##ggsave(`barplot1.png`, width=7, height=10)
```

You can insert any image. 

![A plot with better aspect ratio]


Add some color.

```{r}
my_cols <- rep("black", nrow(candy))
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
my_cols
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) +
  labs(x="Win Percent", y=NULL)
```

> Q17. What is the worst ranked chocolate candy?

Sixlets 


> Q18. What is the best ranked fruity candy?

Starburst 


# Taking a look at pricepercent 

```{r}
candy$pricepercent
```

To see what is a good candy in terms of winpercent and pricepercent, we can make a plot of winpercent vs the pricepercent variable and then see the best candy for the least amount of money.

```{r}
ggplot(candy) + 
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text()

```

To avoid the overplotting of all these labels, we can use an add on package called ggrepl

```{r}
library(ggrepel)

ggplot(candy) + 
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel()
```

Play with the max.overlaps parameter to `geom_text_repel()`

```{r}
# Too hard to see pink 
my_cols[as.logical(candy$fruity)] = "red"

ggplot(candy) + 
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(max.overlaps = 5, col=my_cols)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

# 5 Exploring the correlation structure

```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```

# On to PCA 

The main function for this is `prcom()` and here we need to scale our data with the `scale=TRUE` argument 

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

Plot my main PCA score with ggplot

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

## loadings plot 

```{r}
loadings <- as.data.frame(pca$rotation)

ggplot(loadings) +
  aes(PC1, reorder(rownames(loadings), PC1)) + 
  geom_col()
```

> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

In the positive direction, fruity, pluribus, and hard are picked up strongly by PC1, which does make sense since the correlation indicates that it makes more sense to find hard fruity candy packed together (pluribus).